\emph{scater:} pre-processing, quality control, normalisation and visualisation of single-cell RNA-seq data in R

### 2016-11-06 response_to_reviewers.md

# Aaron's comments

Sweet! Here’s my comments, though keep in mind that this is after about 30 hours w/o sleep, so several grains of salt required:

> Specific comments:
> Major:
> - How do the wrapper functions for running kallisto and Salmon in scater deal with UMI data?  e.g. I don’t believe the current implementation of Salmon has built-in functionality for handle scRNA-seq data with UMIs?

Um. They don’t, I guess?

> - Could the authors add an assessment of how long it takes to perform basic tasks depending on number of genes/cells? With new droplet technologies to sort cells, will R be the best tool to analyze potentially 100s of 1000s of cells?

Point out C++ extension to allow memory- and time-efficient calculations of QC metrics, manipulations, etc. Scales to Zeisel data set (3000 cells, 10000 genes post-filtering) on a desktop machine (8 GB RAM). Larger data sets inevitably require HPC resources; perhaps can get more savings with sparse matrix representations, but this is not easily compatible with downstream methods (e.g., t-SNE, PCA) requiring the full matrix. So just suck it up and fork out some cash.

> - Are there functions to extract the highly variable (or expressed) genes? There are options to visualize most highly expressed features and dim reduction plots that use highly variable genes, but how to extract?  Also, how would you decide what set of genes to look at (e.g. DE genes, highly expressed or highly variable genes) depending on if you were interested in rare cell types versus identifying common major cell types?  Is there an option to control this in the data visualization tools?

Extended in scran - deviation-from-median (DM) method in Kolodziejczyk et al., CV2-based method (technicalCV2) by Brennecke et al., and Lun et al.’s variance of log-expression based methods. HVG selection can be a complicated procedure, so plotPCA/plotTSNE just uses sample variance as a quick-and-dirty default. However, selection based on these more complex methods can be performed by specifying the detected HVGs in feature_set=, allowing consideration of technical contributions to the variance, fitting of the mean-variance trend, rigorous hypothesis testing for HVGs, etc. See example from workflow for how this gets accommodated.

Also planning on testing and implementing the Gini index (Giniclust, can’t remember the paper) as a HVG selection option in scran; focuses on genes that are upregulated in rare cell types, to highlight these cells upon visualisation.

> - For normalization, does it make sense to use bioC packages such as svaseq where if you are trying to discover biological conditions (e.g. novel cell types) from scRNA-seq data, you could be removing the interesting biological variation in the latent factors? What are the author’s opinions of using spike-ins or control genes? Could the authors provide a discussion on this topic or at least refer the reader to other papers to help guide them?

Yes, you could potentially be removing interesting biological variation in the latent variables, in cases where discovery of new and interesting things is the main aim. Perhaps a note about this would be prudent.

Also, see F1000 workflow for discussion of benefits of spike-in normalisation vs non-DE normalisation. Blah blah blah.

> - How did the authors decide on the selection of variable genes? Is there a reason the authors applied the specific normalization of FPKMs to identify the variable genes instead of something like the distance-to-median approach in Kolodziejczyk et al. (2015) Cell Stem Cell. I would like to see a larger discussion of this and I would like the authors to consider other approaches of identifying highly variable genes and how this affects their downstream clustering results?

We used variance of log-values, I believe, in selecting the HVGs within plotPCA, etc.? This has several advantages over CV2-based methods:

1) plotPCA/plotTSNE use log-expression values, so HVGs based on high variance of logged values -> high Euclidean distances, maximise log-fold change and discrimination between cells.
2) log-transformation protect against genes with outlier expression patterns, i.e., very strong expression in only one cell.
3) logged counts are more normal-ish than the raw counts themselves, making it easier to construct tests for significance of HVG’ness.

In general, the quantitative differences between HVG selection methods are modest. Strong HVGs distinguishing cell types get picked up no matter what you do, so you should be able to see separate clusters. It is likely that differences occur in visualisation of more subtle features, e.g., rare cell types (CV2 is better, as it selects outlier-like genes) or weak clusters (where log might be better if it selects more discriminatory genes).

Note, the fact that we’re using FPKMs is irrelevant here, as the length scaling factor cancels out when computing the sample variance within a gene. No mean-variance trend is fitted so the absolute values don’t matter.

> - What if researchers have two types of data (e.g. scRNA-Seq and DNA methylation in scMT-seq https://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-0950-z). How does the SCEset container handle this?  Are there different assays that would allow multiple types of information to be held in the same object?

Not well, as it's outside the design remit of the single cell expression set. But Vince Carey’s MultiAssay object (or something like that, can’t remember exact name) can be used.

> - Stegle et al. 2015 noted there is a big 3’ bias in scRNA-Seq protocols (particularly for the SMART-Seq protocol used by Fluidigm). Therefore, normalizing for transcript length (e.g. FPKM) can lead to an underestimate of expression in long transcripts and over expression in short transcripts. I would like to see the authors create a data visualization function in scater that would allow a user to check this in their own data.

Sure, I guess.

> Minor:
> - Pg 3, line 26: could the authors include new salmon reference listed in bioRxiv (http://biorxiv.org/content/early/2016/08/30/021592)

Yeah, fine, whatever. At least we know who this reviewer is likely to be, then. Who else would care?

> Reviewer: 2
>
> Comments to the Author
> The work by McCarthy and colleagues describes an R package to pre-process and QC single-cell RNA-seq data. The authors did a very nice job with documentation at all levels: manuscript, supplements and especially with the package vignette/case study, which was also provided as supplementary information.
>
> The package itself does not introduce any novel method to QC or normalize single-cell RNA-seq. Rather, it aims at becoming a one-stop solution for loading and organizing single-cell data, which can then be fed to existing methods. While the absence of novel analytical strategies diminishes the scope of the work and may somewhat mitigate its impact, I note that the design choices behind this package, particularly the data structures, are very elegant and polished, and fit well with the R/Bioconductor universe. Visualizations are useful, appealing and pleasant. A regular R user will immediately recognize the quality of this package, which therefore has the potential to become "the" choice for single-cell RNA-seq QC, pretty much like edgeR/DESeq are for differential gene expression.
>
> My points are minor and optional:
>
> - the authors may consider allocating more text to outlier identification. They provide a reference, but it may be worth summarizing briefly how outliers are identified. The metrics used for outlier detection are listed in the package vignette, but I feel this is important information that should go in the main text.

Okay…. are these the QC outliers detected with the PCA-based method in mvoutlier? Guess we could talk about it a bit more?

> - some packages already provide routines for batch correction. This is the case, for instance, of scde. Indeed, scde requires very little upstream pre-processing, mostly just filtering out outliers. So there is no one-size-fits-all workflow. In this regard, I wonder whether the authors might further elaborate on Figure 1 and Supplementary Figure 3 by providing schematics of the most common workflows for the most popular tools, highlighting which portions would be scater-specific and which would be third-party tool-specific?

I’m pretty sure scde requires a bit more work… at the very least, QC on genes and cells isn’t so simple as to be fobbed off in a single sentence. Probably mention the inputs to each of these tools; methods require a count matrix (scde, edgeR) or expression matrix (monocle) or log-expression matrix (plotTSNE, PCA, etc.). Last lot will require batch correction as batches cannot be modelled in the methods themselves.

scde appears to provide one function "clean.counts" for "QC" before running its functions, but this function will only filter cells by minimum library size and genes by minimum total reads and minimum number of cells a gene should be seen in. Thus, preparing a dataset with scater upstream of using scde offers many more possibilities for QC of both cells and genes. scde indeed offers the option to include a known batch effect in its DE models - scater again offers many options for exploring batch effects that should prove useful upstream of modelling with scde.

*Can very easily show some possible workflows with an extra Supp Fig for, for example DE with scde, DE with edgeR, clustering with SC3, ...*

> - the authors did a good job highlighting the possible influence of cell cycle on measurements. I wonder whether the effect of cell cycle should be also mentioned in the latent variable section of the case study. It is not infrequent that a treatment significantly alters the cell cycle, partially masking the phenotype of interest. This can be explicitly investigated by using cell cycle genes to drive PCA, for example, as the authors pointed out. However, there may be cases where the user did not suspect an involvement of cell cycle, but latent variable analysis may reveal it, prompting a more directed investigation.

Fine. Can combined with cyclone (implemented in scran) to determine if levels of the latent variable are associated with cell cycle classification.

> - I wonder whether scater might provide some simple summary stats regarding the distribution of genes. Roughly, genes may show 4 types of distributions: 1) noise 2) zero inflated 3) bimodal 4) log normal. It would be nice if scater provided related information, perhaps upon GMM fitting, such as the number of genes in each class for every sample, and then some basics such as the median expression and standard deviation of log normally distributed genes etc.

Sounds like a pain, not sure what it would tell you. Doesn’t tell you anything about QC - is having lots of bimodal genes good or bad? Who knows? We already get zero-inflation diagnostics from drop-out rate plots. Similarly, it doesn’t tell you much about the interesting genes - log-normally distributed genes can be high or low variance, in which case it is interesting and not interesting respectively. Finally, it’s a pain to draw a clear boundary between noise and zero-inflation (as large dispersions will automatically have a point mass at zero), bimodal and zero inflation (probably the same, if the mean of the non-zero component is large enough), and bimodal and log-normal (test for bimodality - can’t remember the name - seems to involve a few iffy assumptions).

> - minor details: how did the authors set the perplexity parameter of tSNE? How many of the top variable genes are used in PCA and dimensionality reduction in general? Are violin plots scaled to take into account that different groups might have different number of cells?

Default perplexity (20?), 500 genes, dunno, respectively.

-------------------

### 2016-11-06 main.tex
Update to main text
Revisions to manuscript for resubmission

\documentclass{bioinfo}
\copyrightyear{2016} \pubyear{2016}

@@

\title[scater package]{\emph{scater:} pre-processing, quality control, normalisation and visualisation of single-cell RNA-seq data in R}
\author[McCarthy \textit{et~al}.]{Davis J.~McCarthy\,$^{\text{\sfb 1,2,5}*}$, Kieran R.~Campbell\,$^{\text{\sfb 2,4}}$, Aaron T.~L.~Lun\,$^{\text{\sfb 6}}$ and Quin F.~Wills\,$^{\text{\sfb 2,3}}$}
\address{$^{\text{\sf 1}}$European Molecular Biology Laboratory - European Bioinformatics Institute (EMBL-EBI), Hinxton CB10 1SD, United Kingdom;\\
$^{\text{\sf 2}}$Wellcome Trust Centre for Human Genetics, University of Oxford,
Roosevelt Drive, Oxford OX3 7BN, United Kingdom;\\
$^{\text{\sf 3}}$Weatherall Institute for Molecular Medicine, University of Oxford, John Radcliffe Hospital, Oxford OX3 9DS, United Kingdom;\\
$^{\text{\sf 4}}$Department of Physiology, Anatomy and Genetics, University of Oxford, South Parks Road, Oxford OX1 3QX, United Kingdom;\\
$^{\text{\sf 5}}$St Vincent's Institute of Medical Research, 41 Victoria Parade, Fitzroy Victoria 3065, Australia; and \\
$^{\text{\sf 6}}$CRUK Cambridge Institute, University of Cambridge, Robinson Way, Cambridge CB2 0RE, United Kingdom.
% $^{\text{\sf 4}}$Department of Statistics, University of Oxford, Oxford, Oxfordshire, UK;
}

@@

Numerous statistical methods and software tools have been published for scRNA-seq data \citep{Guo2015-wx,Kharchenko2014-rx,Finak2015-rd,Delmans2016-qc,Angerer2015-sw,Kiselev2016-fu,Julia2015-jt,Trapnell2014-gj}. However, all of these assume that quality control and normalisation have already been applied. Fewer methods are available in the literature to perform these basic steps in scRNA-seq data processing \citep{Ilicic2016-dm}. This issue is exacerbated by the diversity of scRNA-seq data sets with respect to the experimental protocol and the biological context of the study, meaning that a single processing pipeline with fixed parameters is unlikely to be universally applicable. Rather, software tools are required that support an interactive approach to analysis. This allows parameters to be fine-tuned for the study at hand in response to any issues diagnosed during data exploration. The provided functionality should also process the data in a statistically rigorous manner and encourage reproducible bioinformatics analyses.

One of the most popular frameworks for interactive analysis is the R programming language, extended for biological data analysis through the Bioconductor project \citep{Huber2015-en}. While Bioconductor packages have been widely used for bulk RNA-seq data, the existing data structures (like the ExpressionSet class) are not sufficient for scRNA-seq data. This is because they do not support data types that are specific to single-cell studies, e.g., cell-cell distance matrices for clustering. For larger studies, this also includes data beyond expression profiles such as intensity values from fluorescence-activated cell sorting, cell imaging data, and information from epigenetic and targeted genotyping assays. Existing methods for processing and applying quality control to scRNA-seq data are similarly inadequate. In particular, current visualisation methods designed for exploratory data analysis of bulk transcriptomic experiments are unsuited to datasets containing hundreds or thousands of cells. The large size of each dataset also favours methods such as \emph{kallisto} \citep{Bray2016-pj} and \emph{Salmon} \citep{Patro2015-jf} for rapidly quantifying gene expression. Extensions to the current computational infrastructure are required to provide appropriate data structures and methods that can accommodate these rich scRNA-seq datasets for integrative analyses of expression and other assay data along with the accompanying metadata.

@@

Further case studies using \emph{scater} on published data, for example from 3000 mouse cortex cells \citep{Zeisel2015-ab} and 1200 cells from early-development mouse embryos \citep{Scialdone2016-oa} are available at \href{http://dx.doi.org/10.5281/zenodo.59897}{dx.doi.org/10.5281/zenodo.59897}.

@@

\begin{figure}[!tpb]%figure1
\centerline{\includegraphics[width=0.48\textwidth]{figures/scater_qc_workflow.pdf}}
\caption{An overview of the \emph{scater} workflow, from raw sequenced reads to a tidy data set ready for higher-level downstream analysis. For step 5, explanatory variables include experimental covariates like batch, cell source and other recorded information, as well as QC metrics computed from the data. Step 6 describes an optional round of normalisation to remove effects of particular explanatory variables from the data. Automated computation of QC metrics and extensive plotting functionality support the workflow.}\label{fig:01}
\end{figure}

@@

The \emph{scater} package is an open-source R package available through     Bioconductor. Key aspects of the code are written in C++ to minimise computational time and memory use. The package builds on many other R packages, including \emph{Biobase} and \emph{BiocGenerics} for core Bioconductor functionality \citep{Huber2015-en}; \emph{destiny} \citep{Angerer2015-sw} and \emph{Rtsne} \citep{Krijthe2015-is} for dimensionality reduction; and \emph{edgeR} \citep{Robinson2010-ky} and \emph{limma} \citep{Ritchie2015-so} for model fitting and statistical analyses. The plotting functionality in the package uses \emph{ggplot2} \citep{Wickham2016-dc}. A full set of dependencies is provided in the Supplementary Materials.

@@

\begin{figure*}[!htpb]%figure3
\centerline{\includegraphics[width=\textwidth]{figures/figure3_alt.pdf}}
\caption{\textbf{Different types of QC plots that can be generated with \emph{scater}.} (\textbf{a}) Cumulative expression plot showing the proportion of the library accounted for by the top 1--500 most highly-expressed features. (\textbf{b}) PCA plot produced using a subset of the QC metrics computed with \emph{scater's} calculateQCMetrics function. (\textbf{c}) Plot of frequency of expression (percentage of cells in which the feature is deemed expressed) against mean expression level across cells. The vertical dotted line shows the median of the gene mean expression levels, and the horizontal dotted line indicates 50\% frequency of expression. (\textbf{d}) Plot of the 20 most highly-expressed features (computed according to the highest total read counts) across all cells in the data set. For each feature, the circle represents the percentage of counts that the gene accounts for computed from counts pooled across all cells. The genes are ordered by this value. The bars for each gene show the percentage of counts accounted for by the gene for each individual cell, providing a visualisation of the distribution across cells. (\textbf{e}) Density plot showing the percentage of variance explained by a set of explanatory variables across all genes. Each individual plot is produced by a single call with either the function plot (\textbf{a}), plotPCA (\textbf{b}) or plotQC (\textbf{c--e}).}\label{fig:03}
\end{figure*}

@@

The plotPCA function implements an approach to automatic outlier
detection using multivariate normal methods applied to the cell-level QC
metrics \citep{Ilicic2016-dm}. Specifically, PCA is applied to the QC metrics
for all cells and a plot is produced to automatically detect outliers in the
higher-dimensional QC metric space (Figure~\ref{fig:03}b). These outliers
correspond to low-quality cells with abnormal library characteristics (e.g.,
low total counts and few expressed genes) that should be removed prior to
downstream analysis. This automated approach is powerful but also somewhat opaque with respect to how outliers are defined, and so complements simpler filtering approaches that apply thresholds to particular QC metrics.

@@

\begin{figure*}[!htpb]%figure4
\centerline{\includegraphics[width=\textwidth]{figures/figure4.pdf}}
\caption{\textbf{Reduced dimension representations of cells and gene expression plots with \emph{scater.}} Plots are shown using all genes (\textbf{a-c}) and cell cycle genes only (\textbf{d-f}) using PCA (\textbf{a,d}), t-SNE (\textbf{b,e}) and diffusion maps (\textbf{c,f}), where each point represents a cell. In the top row (\textbf{a-c}), points are coloured by patient of origin, sized by total features (number of genes with detectable expression) and the shape indicates the C1 machine used to process the cells. In the second row (\textbf{d-f}), points are coloured by the expression of \emph{CCND2} (a gene associated with the G1/S phase transition of the cell cycle) in each cell. With the plotExpression function, gene expression can be plotted against any cell metadata variables or the expression of another gene---here, expression for the CD86, IGH44 and IGHV4-34 genes in each cell is plotted against the patient of origin (\textbf{g}). The function automatically detects whether the x-axis variable is categorical or continuous and plots the data accordingly, with x-axis values ``jittered'' to avoid excessive overplotting of points with the same x coordinate.}\label{fig:04}
\end{figure*}

@@

Scaling normalisation is typically required in RNA-seq data analysis to remove biases caused by differences in sequencing depth, capture efficiency or composition effects between samples. Frequently used methods for scaling normalization include the trimmed mean of M-values \citep{Robinson2010-dy}, relative log-expression \citep{Anders2010-kv} and upper-quartile methods \citep{Bullard2010-ui}, all of which are available for use in \emph{scater}. In addition, \emph{scater} is tightly integrated with the \emph{scran} package that implements a method utilising cell pooling and deconvolution to compute size factors better suited to scRNA-seq data \citep{Lun2016-sk}.

@@

In addition to removing known batch effects, it can be important for
large data sets to identify (potentially unknown) sources of unwanted variation
\citep{Leek2010-nq,Hicks2015-qy,Tung2016-jy,Bacher2016-ay,Grun2015-xi}.
\emph{scater} is compatible with existing methods such as \emph{svaseq}
\citep{Leek2007-rg,Leek2014-nu} and \emph{RUVSeq} \citep{Risso2014-np} to
identify and remove these unwanted sources of variation, and the
removeBatchEffect method in the \emph{limma} package \citep{Ritchie2015-so}
to account for known batch effects. Here, just removing the first latent
variable identified by the RUVs method from \emph{RUVSeq} is sufficient to
remove the machine effect, as the PCA plot now separates cells by patient
rather than C1 machine (Figure~\ref{fig:05}c).

@@

We emphasise that it is generally preferable to incorporate batch effects or latent variables into statistical models used for inference. Where this is not possible (e.g., for visualisation), directly regressing out these uninteresting factors is required to obtain "corrected" expression values for further analysis.

\begin{figure*}[!tpb]%figure5
\centerline{\includegraphics[width=\textwidth]{figures/figure5.pdf}}
\caption{\textbf{Figure 5: Normalisation and batch correction with
\emph{scater}.} Principal component analysis plots showing cell
structure in the first two PCA dimensions using various normalisation
methods methods that can be easily applied in \emph{scater}, including
endogenous size-factor normalisation using methods from the \emph{scran}
package (\textbf{a}); expression residuals after applying size-factor
normalisation and regressing out known, unwanted sources of variation
(\textbf{b}); and removal of one hidden factor identified using
the RUVs method from the \emph{RUV} package (\textbf{c}). In all plots,
the colour of points is determined by the patient from which cells were
obtained, shape is determined by the C1 machine used to process the
cells and size reflects the total number of genes with detectable
expression in the cell.}\label{fig:05}
\end{figure*}

@@

As part of the R/Bioconductor ecosystem, \emph{scater} can be easily
integrated with other software for scRNA-seq data analysis (Supplementary Figure~3). Because the
SCESet class builds on existing Bioconductor data structures, most
Bioconductor packages for expression analyses are able to operate
seamlessly with SCESet objects. Tools that can integrate easily with
\emph{scater} include many options for data normalisation
\citep{Lun2016-sk,Vallejos2015-ww,Ding2015-jv}, differential expression analysis
\citep{Vallejos2016-qy,Trapnell2014-gj,Finak2015-rd,Vu2016-sk,Kharchenko2014-rx,Korthauer2015-wg,Andrews2016-at}, heterogeneous gene expression analyses
\citep{Vallejos2015-ww}, clustering
\citep{Kiselev2016-fu,Guo2015-wx,Fan2016-wh,Grun2015-fy}, latent or hidden
variable analysis
\citep{Leek2014-nu,Risso2014-np,Stegle2012-is,Chikina2015-lq},
cell cycle phase identification \citep{Scialdone2015-gj} and pseudotime
computation \citep{Trapnell2014-gj,Angerer2015-sw,Julia2015-jt,Campbell2016-es,Campbell2015-nj,Haghverdi2016-is}. The \emph{scater} package bridges
the gap between raw reads and these downstream analysis tools by providing the
pre-processing, QC, visualisation and normalisation methods and a data
structure combining multiple data modalities and metadata necessary for
convenient, robust and reproducible analyses of scRNA-seq data.

@@

This work was supported by the National Health and Medical Research Council of Australia [Early Career Fellowship to D.J.M.], by core funding from the European Molecular Biology Laboratory [D.J.M.], core funding from Cancer Research UK [A17197 to A.T.L.L.), the United Kingdom Medical Research Council [studentship to K.R.C.) and the Oxford Single Cell Biology Consortium [Q.F.W.].\vspace*{-12pt}

>>>>

\documentclass{bioinfo}
\newcommand{\revised}[1]{\textcolor{red}{#1}}

\copyrightyear{2016} \pubyear{2016}

@@

\title[scater scRNA-seq package]{\emph{scater:} pre-processing, quality control, normalisation and visualisation of single-cell RNA-seq data in R}
\author[McCarthy \textit{et~al}.~2016]{Davis J.~McCarthy\,$^{\text{\sfb 1,2,5}*}$, Kieran R.~Campbell\,$^{\text{\sfb 2,4}}$, Aaron T.~L.~Lun\,$^{\text{\sfb 6}}$ and Quin F.~Wills\,$^{\text{\sfb 2,3}}$}
\address{$^{\text{\sf 1}}$European Molecular Biology Laboratory - European Bioinformatics Institute (EMBL-EBI), Hinxton CB10 1SD, United Kingdom;\\
$^{\text{\sf 2}}$Wellcome Trust Centre for Human Genetics, University of Oxford,
Roosevelt Drive, Oxford OX3 7BN, United Kingdom;\\
$^{\text{\sf 3}}$Weatherall Institute for Molecular Medicine, University of Oxford, John Radcliffe Hospital, Oxford OX3 9DS, United Kingdom;\\
$^{\text{\sf 4}}$Department of Physiology, Anatomy and Genetics, University of Oxford, South Parks Road, Oxford OX1 3QX, United Kingdom;\\
$^{\text{\sf 5}}$St Vincent's Institute of Medical Research, 41 Victoria Parade, Fitzroy Victoria 3065, Australia; and \\
$^{\text{\sf 6}}$CRUK Cambridge Institute, University of Cambridge, Robinson Way, Cambridge CB2 0RE, United Kingdom.
% $^{\text{\sf 4}}$Department of Statistics, University of Oxford, Oxford, Oxfordshire, UK;
}

@@

Numerous statistical methods and software tools have been published for scRNA-seq data \citep{Guo2015-wx,Kharchenko2014-rx,Finak2015-rd,Angerer2015-sw,Julia2015-jt,Trapnell2014-gj}. However, all of these assume that quality control and normalisation have already been applied. Fewer methods are available in the literature to perform these basic steps in scRNA-seq data processing \citep{Ilicic2016-dm}. This issue is exacerbated by the diversity of scRNA-seq data sets with respect to the experimental protocol and the biological context of the study, meaning that a single processing pipeline with fixed parameters is unlikely to be universally applicable. Rather, software tools are required that support an interactive approach to analysis. This allows parameters to be fine-tuned for the study at hand in response to any issues diagnosed during data exploration. The provided functionality should also process the data in a statistically rigorous manner and encourage reproducible bioinformatics analyses.

One of the most popular frameworks for interactive analysis is the R programming language, extended for biological data analysis through the Bioconductor project \citep{Huber2015-en}. While Bioconductor packages have been widely used for bulk RNA-seq data, the existing data structures (like the ExpressionSet class) are not sufficient for scRNA-seq data. This is because they do not support data types that are specific to single-cell studies, e.g., cell-cell distance matrices for clustering. For larger studies, this also includes data beyond expression profiles such as intensity values from fluorescence-activated cell sorting, cell imaging data, and information from epigenetic and targeted genotyping assays. Existing methods for processing and applying quality control to scRNA-seq data are similarly inadequate. In particular, current visualisation methods designed for exploratory data analysis of bulk transcriptomic experiments are unsuited to datasets containing hundreds or thousands of cells. The large size of each dataset also favours methods such as \emph{kallisto} \citep{Bray2016-pj} and \emph{Salmon} \revised{\citep{Patro2016-rs}} for rapidly quantifying gene expression. Extensions to the current computational infrastructure are required to provide appropriate data structures and methods that can accommodate these rich scRNA-seq datasets for integrative analyses of expression and other assay data along with the accompanying metadata.

@@

Further case studies using \emph{scater} on published data, for example from 3000 mouse cortex cells \citep{Zeisel2015-ab} and 1200 cells from early-development mouse embryos \citep{Scialdone2016-oa} are available at \href{http://dx.doi.org/10.5281/zenodo.59897}{dx.doi.org/10.5281/zenodo.59897}. All materials required to reproduce the results presented in this paper are available at \href{http://dx.doi.org/10.5281/zenodo.60139}{dx.doi.org/10.5281/zenodo.60139}.

@@

\begin{figure}[!tpb]%figure1
\centerline{\includegraphics[width=0.48\textwidth]{scater_qc_workflow.eps}}
\caption{An overview of the \emph{scater} workflow, from raw sequenced reads to a tidy data set ready for higher-level downstream analysis. For step 5, explanatory variables include experimental covariates like batch, cell source and other recorded information, as well as QC metrics computed from the data. Step 6 describes an optional round of normalisation to remove effects of particular explanatory variables from the data. Automated computation of QC metrics and extensive plotting functionality support the workflow.}\label{fig:01}
\end{figure}

@@

The \emph{scater} package is an open-source R package available through     Bioconductor. Key aspects of the code are written in C++ to minimise computational time and memory use\revised{, and the package scales well to large datasets. For example, consider the \citet{Macosko2015-vt} dataset, which contains more than 44,000 cells. The core scater functions to create an SCESet object and calculate QC metrics took approximately two minutes to complete on an early 2015 MacBook Pro laptop with 2.9 GHz Intel Core i5 processor and 16 Gb of RAM. Subsetting the SCESet object takes only a few seconds, and producing a PCA plot with the plotPCA function takes less than a minute.}

The package builds on many other R packages, including \emph{Biobase} and \emph{BiocGenerics} for core Bioconductor functionality \citep{Huber2015-en}; \emph{destiny} \citep{Angerer2015-sw} and \emph{Rtsne} for dimensionality reduction; and \emph{edgeR} \citep{Robinson2010-ky} and \emph{limma} \citep{Ritchie2015-so} for model fitting and statistical analyses. The plotting functionality in the package uses \emph{ggplot2} \citep{Wickham2016-dc}. A full set of dependencies is provided in the Supplementary Materials.

@@

\begin{figure*}[!htpb]%figure3
\centerline{\includegraphics[width=\textwidth]{figure3_alt.eps}}
\caption{\textbf{Different types of QC plots that can be generated with \emph{scater}.} (\textbf{a}) Cumulative expression plot showing the proportion of the library accounted for by the top 1--500 most highly-expressed features. (\textbf{b}) PCA plot produced using a subset of the QC metrics computed with \emph{scater's} calculateQCMetrics function. (\textbf{c}) Plot of frequency of expression (percentage of cells in which the feature is deemed expressed) against mean expression level across cells. The vertical dotted line shows the median of the gene mean expression levels, and the horizontal dotted line indicates 50\% frequency of expression. (\textbf{d}) Plot of the 20 most highly-expressed features (computed according to the highest total read counts) across all cells in the data set. For each feature, the circle represents the percentage of counts that the gene accounts for computed from counts pooled across all cells. The genes are ordered by this value. The bars for each gene show the percentage of counts accounted for by the gene for each individual cell, providing a visualisation of the distribution across cells. (\textbf{e}) Density plot showing the percentage of variance explained by a set of explanatory variables across all genes. Each individual plot is produced by a single call with either the function plot (\textbf{a}), plotPCA (\textbf{b}) or plotQC (\textbf{c--e}).}\label{fig:03}
\end{figure*}

@@

\revised{The plotPCA function implements an approach to automatic outlier
detection using multivariate normal methods applied to the cell-level QC
metrics \citep{Ilicic2016-dm} and a PCA plot computed from QC metrics is produced showing detected outliers (Figure~\ref{fig:03}b). Briefly, semi-robust principal components are computed on robustly sphered QC metrics data to determine distances for each cell and weights for outlier detection, using the pcout function from the \emph{mvoutlier} R package, which implements the fast algorithm of \citet{Filzmoser2008-sy}. By default, the following QC metrics are used for automatic outlier detection: the percentage of counts from the top 100 features, the total number of features with detectable expression, the percentage of counts from control features, the number of detected feature controls, the log-scaled counts from endogenous features and log-transformed counts from feature controls are used. The user can specify QC metrics or other cell metadata variables to use for outlier detection with the "selected\_variables" argument to plotPCA. Detected} outliers
correspond to low-quality cells with abnormal library characteristics (e.g.,
low total counts and few expressed genes) that should be removed prior to
downstream analysis. This automated approach is powerful but also somewhat opaque with respect to how outliers are defined, and so complements simpler filtering approaches that apply thresholds to particular QC metrics.

@@

\begin{figure*}[!htpb]%figure4
\centerline{\includegraphics[width=\textwidth]{figure4.eps}}
\caption{\textbf{Reduced dimension representations of cells and gene expression plots with \emph{scater.}} Plots are shown using all genes (\textbf{a-c}) and cell cycle genes only (\textbf{d-f}) using PCA (\textbf{a,d}), t-SNE (\textbf{b,e}) and diffusion maps (\textbf{c,f}), where each point represents a cell. In the top row (\textbf{a-c}), points are coloured by patient of origin, sized by total features (number of genes with detectable expression) and the shape indicates the C1 machine used to process the cells. In the second row (\textbf{d-f}), points are coloured by the expression of \emph{CCND2} (a gene associated with the G1/S phase transition of the cell cycle) in each cell. With the plotExpression function, gene expression can be plotted against any cell metadata variables or the expression of another gene---here, expression for the CD86, IGH44 and IGHV4-34 genes in each cell is plotted against the patient of origin (\textbf{g}). The function automatically detects whether the x-axis variable is categorical or continuous and plots the data accordingly, with x-axis values ``jittered'' to avoid excessive overplotting of points with the same x coordinate.}\label{fig:04}
\end{figure*}

@@

Scaling normalisation is typically required in RNA-seq data analysis to remove biases caused by differences in sequencing depth, capture efficiency or composition effects between samples. Frequently used methods for scaling normalization include the trimmed mean of M-values \citep{Robinson2010-dy}, relative log-expression \citep{Anders2010-kv} and upper-quartile methods \citep{Bullard2010-ui}, all of which are available for use in \emph{scater}. In addition, \emph{scater} is tightly integrated with the \emph{scran} package that implements a method utilising cell pooling and deconvolution to compute size factors better suited to scRNA-seq data \citep{Lun2016-sk}. \revised{\citet{Lun2016-uj} also offers further discussion of the respective benefits and drawbacks of spike-in normalisation and non-DE normalisation.}

@@

In addition to removing known batch effects, it can be important for
large data sets to identify (potentially unknown) sources of unwanted variation
\citep{Leek2010-nq,Hicks2015-qy,Grun2015-xi}.
\emph{scater} is compatible with existing methods such as \emph{svaseq}
\citep{Leek2007-rg,Leek2014-nu} and \emph{RUVSeq} \citep{Risso2014-np} to
identify and remove these unwanted sources of variation, and the
removeBatchEffect method in the \emph{limma} package \citep{Ritchie2015-so}
to account for known batch effects. Here, just removing the first latent
variable identified by the RUVs method from \emph{RUVSeq} is sufficient to
remove the machine effect, as the PCA plot now separates cells by patient
rather than C1 machine (Figure~\ref{fig:05}c). \revised{More targeted applications of these methods can be used to remove specific effects, for example, by identifying latent factors from cell cycle genes to remove the cell cycle effect.}

@@

We emphasise that it is generally preferable to incorporate batch effects or latent variables into statistical models used for inference. Where this is not possible (e.g., for visualisation), directly regressing out these uninteresting factors is required to obtain "corrected" expression values for further analysis. \revised{A general risk of removing latent factors is that interesting biological variation may be removed along with the presumed unwanted variation. Users should therefore apply such methods with appropriate caution, particularly when an analysis aims to discover biological conditions, such as new cell types.}

\begin{figure*}[!tpb]%figure5
\centerline{\includegraphics[width=\textwidth]{figure5.eps}}
\caption{\textbf{Normalisation and batch correction with
\emph{scater}.} Principal component analysis plots showing cell
structure in the first two PCA dimensions using various normalisation
methods that can be easily applied in \emph{scater}, including
endogenous size-factor normalisation using methods from the \emph{scran}
package (\textbf{a}); expression residuals after applying size-factor
normalisation and regressing out known, unwanted sources of variation
(\textbf{b}); and removal of one hidden factor identified using
the RUVs method from the \emph{RUV} package (\textbf{c}). In all plots,
the colour of points is determined by the patient from which cells were
obtained, shape is determined by the C1 machine used to process the
cells and size reflects the total number of genes with detectable
expression in the cell.}\label{fig:05}
\end{figure*}

@@

As part of the R/Bioconductor ecosystem, \emph{scater} can be easily
integrated with other software for scRNA-seq data analysis (Supplementary Figure~3). Because the
SCESet class builds on existing Bioconductor data structures, most
Bioconductor packages for expression analyses are able to operate
seamlessly with SCESet objects. Tools that can integrate easily with
\emph{scater} include many options for data normalisation
\citep{Lun2016-sk,Vallejos2015-ww,Ding2015-jv}, differential expression analysis
\citep{Vallejos2016-qy,Trapnell2014-gj,Finak2015-rd,Vu2016-sk,Kharchenko2014-rx,Korthauer2015-wg,Andrews2016-at}, heterogeneous gene expression analyses
\citep{Vallejos2015-ww}, clustering
\citep{Kiselev2016-fu,Guo2015-wx,Fan2016-wh,Grun2015-fy}, latent or hidden
variable analysis
\citep{Leek2014-nu,Risso2014-np,Stegle2012-is,Chikina2015-lq},
cell cycle phase identification \citep{Scialdone2015-gj} and pseudotime
computation \citep{Trapnell2014-gj,Angerer2015-sw,Julia2015-jt,Campbell2016-es,Haghverdi2016-is}. The \emph{scater} package bridges
the gap between raw reads and these downstream analysis tools by providing the
pre-processing, QC, visualisation and normalisation methods and a data
structure combining multiple data modalities and metadata necessary for
convenient, robust and reproducible analyses of scRNA-seq data \revised{(see Supplementary Material for discussion of entry points to several third party tools from \emph{scater}).}

@@

This work was supported by the National Health and Medical Research Council of Australia [APP1112681 to D.J.M.], by core funding from the European Molecular Biology Laboratory [D.J.M.], core funding from Cancer Research UK [A17197 to A.T.L.L.), the United Kingdom Medical Research Council [studentship to K.R.C.) and the Oxford Single Cell Biology Consortium [Q.F.W.].\vspace*{-12pt}
